Covariate analysis of MSSM reprocessed counts (with TMM normalisation and covariates adjusted)

### 2017-03-20 code/Rmd/MSSM_REPROCESSED_TMM.Rmd
Added markdown to process MSSM reprocessed RNASeq with TMM and SVA normalaisation

---
title: "Covariate analysis of MSSM reprocessed counts (with CQN normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED.Rmd",
                                 parentId = "syn7554308",
                                 entityName = 'MSSM Reprocessed RNASeq (Normalised)')
```

@@

thisFileName <- 'MSSM_REPROCESSED.Rmd'

@@

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=5, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$individualIdentifier)

  # Re-estimate voom weights
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                              design=DM1$design, plot=F, 
                              block = COVARIATES$individualIdentifier, 
                              correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights,
                            block = COVARIATES$individualIdentifier, 
                            correlation = correlation$cor)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
FIT = lmFit(VOOM.WEIGHTS)

@@

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=cbind(MODEL1, SURR.VAR), plot=F)

# Calculate correlation values of random effects
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = cbind(MODEL1, SURR.VAR), plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

@@

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=DESIGN.NET, plot=F)

# Calculate correlation values of random effects
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = DESIGN.NET, plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

@@

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "MSSM Reprocessed RNASeq (Normalised)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store logCPM
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('MSSM_FP_STG_PHG_IFG_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

@@

# Store differential expression results
write.table(DE, file = 'MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', 
                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

>>>>

---
title: "Covariate analysis of MSSM reprocessed counts (with TMM normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED_TMM.Rmd",
                                 parentId = "syn7554308",
                                 entityName = 'MSSM Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)')
```

@@

thisFileName <- 'MSSM_REPROCESSED_TMM.Rmd'

@@

### Normalisation (with NULL)
Initial normalisation usign TMM and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Normalise using TMM
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-normalise using TMM
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(VOOM.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=5, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

@@

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$individualIdentifier)

  # Re-estimate voom weights
  VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, 
                              design=DM1$design, plot=F, 
                              block = COVARIATES$individualIdentifier, 
                              correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                VOOM.GENE_EXPRESSION$E)

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(TMM.GENE_EXPRESSION,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
FIT = lmFit(VOOM.WEIGHTS)

@@

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=cbind(MODEL1, SURR.VAR), plot=F)

# Calculate correlation values of random effects
correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(TMM.GENE_EXPRESSION, 
                        design = cbind(MODEL1, SURR.VAR), plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  VOOM.SVA.WEIGHTS$E)
```

@@

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(TMM.GENE_EXPRESSION, 
                        design=DESIGN.NET, plot=F)

# Calculate correlation values of random effects
correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)

# Re-estimate voom weights
VOOM.NET.WEIGHTS = voom(TMM.GENE_EXPRESSION, 
                        design = DESIGN.NET, plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  VOOM.NET.WEIGHTS$E)
```

@@

  normalizationType	= 'TMM'
)

# Code
CODE <- Folder(name = "MSSM Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store logCPM
VOOM.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

@@

# Store differential expression results
write.table(DE, file = 'MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MSSM_FP_STG_PHG_IFG_DiffExpression.tsv', 
                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
